Drought indices of water demand and supply and their impact on LULCC in Chile from ERA5-Land and MODIS

### 2024-01-25 article/paper_drought_Chile2023.qmd
actualizar jueves

@Vicente-Serrano2021, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950. Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction. @Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity"_. In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover macroclasses (see @sec-persistence_mask), to avoid misleading conclusions that could be related to the increase in water demand due to LULCC. Our results show a contrasting perspective. There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the Mega Drought [@Garreaud2017]. Despite we used the persistance mask for vegetations's trend analysis, the cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types. These changes are associated with a decrease in water demand from vegetation. On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales and consequently impact the hydrological system. We claim that what occurred in central Chile defies findings made at the global level [@Vicente-Serrano2021;@Kogan2020], demonstrating that a constant decrease in water supply rather than an increase in water demand (i.e., irrigated crops) is the main cause of the hydrological drought. Finally, central Chile has shown a diminishment in vegetation productivity for all macroclasses, mainly attributed to variation in water supply, i.e., precipitation, which could be strengthened by an increase in water demand by, for example, irrigated crops.

1.-  Respecto a lo que indica @Vicente-Serrano2022, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas) que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en Chile?

@Chen2022: Assessing the impact of drought-land cover change on global vegetation greenness and productivity

@Kogan2020: muestra que la sequía no se ha intensificado ni expandido a nivel global, por 38 años.

@@

Extreme drought conditions are an important driver of tree mortality, as shown by @Senf2020 in Europe. However, we found that forest is the type of landcover macroclass less affected by variation in drought indices, being the most resistant landcover class to drought. Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly. Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance. @Xiao2023, using Vegetation Optical Depth (VOD), kNDVI, and EVI, evaluates ecosystem resistance, showing that, compared to croplands, ecosystems with a higher proportion of forests are more resilient to severe droughts. They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023]. Despite our insights, @Venegas2022, who studied species of sclerophyllous populations, indicates that the trees experienced a general decline in growth, and the natural dynamics of their forests may have been altered, mainly due to the cumulative effects of the unprecedented drought (i.e., hydrological drought). Nonetheless, the species of sclerophyllous forest were not the predominant surface within the forest macroclass used in this study. (REVISAR ESTO, LA ultima parte lo que pasó en el bvosque esclerofilo fue el 2020, año extremo)

2.- Sobre los tipos de land cover más afectados por los indicadores de sequía. Asociación con el matorral chileno [@Fuentes2021]. Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur (@Miranda2023)

@Fathi-Taperasht2022: muestra que bosques son más resilentes a la sequía en India.

@Wu2024: evalua la perdida de vegetación en diferentes tipos de uso de suelo. Muestra que bosques son más resilentes y cultivos y pastizales, los menos.

@Jiang2020: impacto de la sequía en landcover en China.

>>>>

@Vicente-Serrano2021, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950. Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction. @Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity"_.

In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover macroclasses (see @sec-persistence_mask), to avoid misleading conclusions that could be related to the increase in water demand due to LULCC. Our results show a contrasting perspective. There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the Mega Drought [@Garreaud2017]. Despite using the persistance mask for vegetation's trend analysis, cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types. These changes are associated with a decrease in water demand from vegetation. Nonetheless, we used the persistant landcover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption. But this effect should be minor compared to the results from ladcover macroclasses.

On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales and consequently impact the hydrological system. We claim that what occurred in central Chile defies findings made at the global level [@Vicente-Serrano2021;@Kogan2020], demonstrating that a constant decrease in water supply rather than an increase in water demand (i.e., irrigated crops) is the main cause of the hydrological drought. Finally, central Chile has shown a diminishment in vegetation productivity for all macroclasses, mainly attributed to variation in water supply, i.e., precipitation, which could be strengthened by an increase in water demand by, for example, an increase in the surface area of irrigated crops.

@@

We found that shrubland, savannas (Chilean matorral), and croplands are the most sensitive to climate conditions. Being most affected by the 12-month soil moisture deficit. In a study in the Yangtze River Basin in China, Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI). They found that cropland was more sensitive to drought than cropland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6). In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016; @Potopova2015; @Dai2020]. Short-term precipitation deficits impact soil water, and thus less water is available for plant growth. However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with hydrological drought (long-term), and in "Centro," an SPI-12 and SPEI-12. Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater. In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared (\ref{tab-corland cover}). What can be explained is that, south of "Centro," predominate forest and grassland, the most resistant types. Also, drought episodes have been less frequent and intense. The drought episodes have had a lower impact on water availability for vegetation.

Extreme drought conditions are an important driver of tree mortality, as shown by @Senf2020 in Europe. However, we found that forest is the type of landcover macroclass less affected by variation in drought indices, being the most resistant landcover class to drought. Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly. Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance. Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023, tests the resistance of ecosystems and finds that ecosystems with more forests are better able to handle severe droughts than croplands. They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].

In contrast to our observations, @Venegas2022, who investigated Cryptocarya alba and Beilschmiedia miersii (both members of the Lauraceae family) inhabiting sclerophyllous forests in Chile, concluded that the overall development of the trees had decelerated, potentially indicating alterations in the inherent dynamics of their respective forests. They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought). Nonetheless, the trees within sclerophyllous forests are highly heterogeneous, and the species analyzed by @Venegas2022 are not the predominant ones. Because of this, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer. Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity. Then, for a study of the dynamics of forest to drought, e.g., sclerophyllous forest, on a local scale using remote sensing, we should use operational products at higher spatial resolutions as provided by the missions Landsat or Sentinel that could capture the heterogenity of the forest type.

## ERA5-Land derived soil moisture and vegetation productivity

### 2024-01-25 article/paper_drought_Chile2023.tex
actualizar jueves

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden]}{\end{tcolorbox}}\fi

@@

\citep{Vicente-Serrano2021}, in a study at the global scale of drought
trends, indicates that there have not been significant trends in
meteorological drought since 1950. Also, state that the increase in
hidrological trend in some parts of the globe (northeast Brazil and the
Mediterranean region) is related to changes in land cover and
specifically to the rapidly increasing irrigated area, which
consequently increases water extraction. \citep{Kogan2020} analyzed the
agricultural drought impact globally and in the main grain producer
countries, finding that \emph{``since 1980, the Earth warming has not
changed the drought area or intensity''}. In our study, we considered
the variation in vegetation productivity in Chile for areas without
changes in land cover macroclasses (see
Section~\ref{sec-persistence_mask}), to avoid misleading conclusions
that could be related to the increase in water demand due to LULCC. Our
results show a contrasting perspective. There has been a significant
trend in the decline of vegetation productivity (zcNDVI) since 2000 for
``Norte Chico'' and ``Centro,'' which has been extreme between 2020 and
2022, seemsly due to an intense hydrological drought due to the
persistance of the Mega Drought \citep{Garreaud2017}. Despite we used
the persistance mask for vegetations's trend analysis, the cropland,
which is the most water-demand type, showed a decrease trend in ``Norte
Chico'' and ``Centro.'' Also, there was an increase in barren land for
both types. These changes are associated with a decrease in water demand
from vegetation. On the other hand, for ``Norte Chico'' and ``Centro,''
our results show a decrease in trends of water supply (SPI and SSI),
which are higher at larger time scales and consequently impact the
hydrological system. We claim that what occurred in central Chile defies
findings made at the global level
\citep{Vicente-Serrano2021, Kogan2020}, demonstrating that a constant
decrease in water supply rather than an increase in water demand (i.e.,
irrigated crops) is the main cause of the hydrological drought. Finally,
central Chile has shown a diminishment in vegetation productivity for
all macroclasses, mainly attributed to variation in water supply, i.e.,
precipitation, which could be strengthened by an increase in water
demand by, for example, irrigated crops.

1.- Respecto a lo que indica \citep{Vicente-Serrano2022}, de que el
aumento en la tendencia en severidad de la sequía (hidrológica) tiene
que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas)
que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en
Chile?

\citep{Chen2022}: Assessing the impact of drought-land cover change on
global vegetation greenness and productivity

\citep{Kogan2020}: muestra que la sequía no se ha intensificado ni
expandido a nivel global, por 38 años.

@@

Extreme drought conditions are an important driver of tree mortality, as
shown by \citep{Senf2020} in Europe. However, we found that forest is
the type of landcover macroclass less affected by variation in drought
indices, being the most resistant landcover class to drought. Supporting
this is \citep{Fathi-Taperasht2022}, who asserts that Indian forests are
the most drought-resistant and recover rapidly. Similarly, the work of
\citep{Wu2024}, who analyzed vegetation loss and recovery in response to
meteorological drought in the humid subtropical Pearl River basin in
China, indicates that forests showed higher drought resistance.
\citep{Xiao2023}, using Vegetation Optical Depth (VOD), kNDVI, and EVI,
evaluates ecosystem resistance, showing that, compared to croplands,
ecosystems with a higher proportion of forests are more resilient to
severe droughts. They attribute the difference to a deeper rooting depth
of trees, a higher water storage capacity, and different water use
strategies between forest and cropland \citep{Xiao2023}. Despite our
insights, \citep{Venegas2022}, who studied sclerophyllous populations,
indicates that the trees experienced a general decline in growth, and
the natural dynamics of their forests may have been altered, mainly due
to the cumulative effects of the unprecedented drought (i.e.,
hydrological drought). Nonetheless, sclerophyllous forest is not the
predominant surface within the forest macroclass used in this study.

2.- Sobre los tipos de land cover más afectados por los indicadores de
sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur
(\citep{Miranda2023})

\citep{Fathi-Taperasht2022}: muestra que bosques son más resilentes a la
sequía en India.

\citep{Wu2024}: evalua la perdida de vegetación en diferentes tipos de
uso de suelo. Muestra que bosques son más resilentes y cultivos y
pastizales, los menos.

\citep{Jiang2020}: impacto de la sequía en landcover en China.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

\citep{Vicente-Serrano2021}, in a study at the global scale of drought
trends, indicates that there have not been significant trends in
meteorological drought since 1950. Also, state that the increase in
hidrological trend in some parts of the globe (northeast Brazil and the
Mediterranean region) is related to changes in land cover and
specifically to the rapidly increasing irrigated area, which
consequently increases water extraction. \citep{Kogan2020} analyzed the
agricultural drought impact globally and in the main grain producer
countries, finding that \emph{``since 1980, the Earth warming has not
changed the drought area or intensity''}.

In our study, we considered the variation in vegetation productivity in
Chile for areas without changes in land cover macroclasses (see
Section~\ref{sec-persistence_mask}), to avoid misleading conclusions
that could be related to the increase in water demand due to LULCC. Our
results show a contrasting perspective. There has been a significant
trend in the decline of vegetation productivity (zcNDVI) since 2000 for
``Norte Chico'' and ``Centro,'' which has been extreme between 2020 and
2022, seemsly due to an intense hydrological drought due to the
persistance of the Mega Drought \citep{Garreaud2017}. Despite using the
persistance mask for vegetation's trend analysis, cropland, which is the
most water-demand type, showed a decrease trend in ``Norte Chico'' and
``Centro.'' Also, there was an increase in barren land for both types.
These changes are associated with a decrease in water demand from
vegetation. Nonetheless, we used the persistant landcover to ensure that
the pixel has the same class; in the case of croplands, it could happen
that some areas had changed crops for others with higher water
consumption. But this effect should be minor compared to the results
from ladcover macroclasses.

On the other hand, for ``Norte Chico'' and ``Centro,'' our results show
a decrease in trends of water supply (SPI and SSI), which are higher at
larger time scales and consequently impact the hydrological system. We
claim that what occurred in central Chile defies findings made at the
global level \citep{Vicente-Serrano2021, Kogan2020}, demonstrating that
a constant decrease in water supply rather than an increase in water
demand (i.e., irrigated crops) is the main cause of the hydrological
drought. Finally, central Chile has shown a diminishment in vegetation
productivity for all macroclasses, mainly attributed to variation in
water supply, i.e., precipitation, which could be strengthened by an
increase in water demand by, for example, an increase in the surface
area of irrigated crops.

@@

We found that shrubland, savannas (Chilean matorral), and croplands are
the most sensitive to climate conditions. Being most affected by the
12-month soil moisture deficit. In a study in the Yangtze River Basin in
China, Jiang2020 analyzed the impact of drought on vegetation using the
SPEI and the Enahanced Vegetation Index (EVI). They found that cropland
was more sensitive to drought than cropland, showing that cropland
responds strongly to short- and medium-term drought (\textless{}
SPEI-6). In our case, the SPEI-12 was the one that most impacted the
croplands in ``Norte Chico'' and ``Centro.'' In general, most studies
show that croplands are most sensitive to short-term drought
(\textless{} SPI-6) \citep{Zambrano2016, Potopova2015, Dai2020}.
Short-term precipitation deficits impact soil water, and thus less water
is available for plant growth. However, we found that in ``Norte
Chico,'' an SPI-36 and SPEI-12 had a higher impact, which are associated
with hydrological drought (long-term), and in ``Centro,'' an SPI-12 and
SPEI-12. Thus, we attribute this impact to the hydrological drought that
has decreased groundwater storage \citep{Taucare2024}, which in turn is
impacted by long-term deficits, and consequently, the vegetation is more
dependent on groundwater. In ``Sur'' and ``Austral,'' the correlations
between drought indices and vegetation productivity decrease, as do the
time scales that reach the maximum r-squared (\ref{tab-corland cover}).
What can be explained is that, south of ``Centro,'' predominate forest
and grassland, the most resistant types. Also, drought episodes have
been less frequent and intense. The drought episodes have had a lower
impact on water availability for vegetation.

Extreme drought conditions are an important driver of tree mortality, as
shown by \citep{Senf2020} in Europe. However, we found that forest is
the type of landcover macroclass less affected by variation in drought
indices, being the most resistant landcover class to drought. Supporting
this is \citep{Fathi-Taperasht2022}, who asserts that Indian forests are
the most drought-resistant and recover rapidly. Similarly, the work of
\citep{Wu2024}, who analyzed vegetation loss and recovery in response to
meteorological drought in the humid subtropical Pearl River basin in
China, indicates that forests showed higher drought resistance. Using
Vegetation Optical Depth (VOD), kNDVI, and EVI, \citep{Xiao2023}, tests
the resistance of ecosystems and finds that ecosystems with more forests
are better able to handle severe droughts than croplands. They attribute
the difference to a deeper rooting depth of trees, a higher water
storage capacity, and different water use strategies between forest and
cropland \citep{Xiao2023}.

In contrast to our observations, \citep{Venegas2022}, who investigated
Cryptocarya alba and Beilschmiedia miersii (both members of the
Lauraceae family) inhabiting sclerophyllous forests in Chile, concluded
that the overall development of the trees had decelerated, potentially
indicating alterations in the inherent dynamics of their respective
forests. They attributed it to the cumulative effects of the
unprecedented drought (i.e., hydrological drought). Nonetheless, the
trees within sclerophyllous forests are highly heterogeneous, and the
species analyzed by \citep{Venegas2022} are not the predominant ones.
Because of this, when we used MOD13A3 with a 1km spatial resolution to
measure vegetation condition, it took the average condition of 1 square
kilometer. Thus, we attribute that forest to being the most resistant to
drought, due to the fact that most of the species comprising it are
highly resilient to water scarcity. Then, for a study of the dynamics of
forest to drought, e.g., sclerophyllous forest, on a local scale using
remote sensing, we should use operational products at higher spatial
resolutions as provided by the missions Landsat or Sentinel that could
capture the heterogenity of the forest type.
